Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-02-27 write-ups/FSE/paper.tex
got everything but the conclusion to fit

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{overview.pdf}
\caption{Cooperative kernels can flexibly resize to let other tasks
(e.g. graphics) run concurrently}
\label{fig:overview}
\end{figure}

@@

\myfiglong{\ref{fig:overview}} illustrates sharing of GPU compute units between a cooperative kernel and a
graphics task. Workgroups 2 and 3 of the cooperative kernel
are killed at an $\offerkill$ to make room for a graphics
task. The workgroups are subsequently restored to the cooperative kernel when workgroup 0 calls $\offerfork$. The order in
which workgroups are killed is important and discussed in
\mysec{\ref{sec:semantics}}. The \emph{gather time} is the time
between resources being requested and the application surrendering them via $\offerkill$. To satisfy soft-real time
constraints, this time should be low; our experimental study
(\mysec{\ref{sec:responsiveness}}) shows that in practice, the
gather-time for our applications is acceptable for a range of graphics
workloads.

@@

\begin{lstlisting}
(*@\label{line:wksteal:kernelfunc}@*)kernel work_stealing(global Task * queues) {
(*@\label{line:wksteal:getgroupid}@*)  int queue_id = get_group_id();
(*@\label{line:wksteal:mainloop}@*)  while (more_work(queues)) {
(*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
    if (t) {
(*@\label{line:wksteal:processtask}@*)      process_task(t, queues, queue_id);
    }
  }
}
\end{lstlisting}
\vspace{-2mm}
\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
%\vspace{-6mm}
\end{figure}

@@

\begin{lstlisting}
kernel graph_app(global graph * g,
                 global nodes * n0, global nodes * n1) {
  int level = 0;
  global nodes * in_nodes = n0;
  global nodes * out_nodes = n1;
  int tid = get_global_id();
  int stride = get_global_size();
(*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
    for (int i = tid; i < nodes.size; i += stride) {
      process_node(g, in_nodes[i], out_nodes, level);
    }
(*@\label{line:graph:swap}@*)    swap(&in_nodes, &out_nodes);
(*@\label{line:graph:gb1}@*)    global_barrier();
(*@\label{line:graph:reset}@*)    reset(out_nodes);
    level++;
(*@\label{line:graph:gb2}@*)    global_barrier();
  }
}
\end{lstlisting}
\vspace{-2mm}
\caption{An OpenCL graph traversal algorithm}\label{fig:graphsearch}
%\vspace{-2mm}
\end{figure}

@@

\begin{lstlisting}
kernel graph_app(global graph *g,
                 global nodes *n0, global nodes *n1) {
(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
  while(in_nodes.size > 0) {
(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
    for (int i = tid; i < nodes.size; i += stride) {
      process_node(g, in_nodes[i], out_nodes, level);
    }
    swap(&in_nodes, &out_nodes);
(*@\label{line:cgraph:resizing1}@*)    resizing_global_barrier();
    reset(out_nodes);
    level++;
(*@\label{line:cgraph:resizing2}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
%\vspace{-2mm}
\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
\vspace{-4mm}
\end{figure}

@@

On receiving a data packet describing a kernel launch $K$, the GPU
scheduler must decide how to schedule $K$. Suppose $K$ requests $N$
workgroups. The scheduler queries the scheduler context.  If there are
at least $N$ available workgroups, $K$ can be scheduled
immediately. Suppose instead that there are only $N_a < N$ available
workgroups, but a cooperative kernel $K_c$ is executing. The scheduler
can use $K_c$'s channel in the scheduler context to command $K_c$ to
provide $N - N_a$ workgroups via $\offerkill$.  Once $N$ workgroups
are available,
% the scheduler initialises a kernel context $\KC$ for $K$, setting
%the number of active workgroups to $N$ and mapping available
%workgroups (located in the scheduler context) to their new $\KC$
%workgroup ids.
the scheduler then sends $N$ workgroups from the available workgroups
to execute kernel $K$
% with
%kernel context $\KC$.
(If the new kernel $K$ is itself a cooperative kernel, the scheduler
would be free to provide $K$ with fewer than $N$ active workgroups
initially.)

@@

\begin{table}[t]
\normalsize
\centering
\begin{tabular}{ l r r r r r r}
App. & barriers & kill & fork & transmit & LoC & inputs\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
%\rowcolor{Gray1}
%bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
\rowcolor{Gray3}
octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
\rowcolor{Gray3}
game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
\end{tabular} \\
%\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
\vspace{-8mm}
\end{table}

@@

Among the GPUs available to us, four met our requirements: Intel HD
520, HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux
drivers for OpenCL suffer from a known defect whereby long-running
kernels lead to unkillable defunct
processes~\cite{amd-defunct-process}.
%
%and simultaneously obtained alarming results on this platform when
%using SVM atomics: modifying a kernel to include a simple host/device
%handshake through SVM slowed down kernel execution by an order of
%magnitude in some cases.  This behaviour, combined with the defunct
%processes issue, lead us to suspect that the support we require for
%cooperative kernels is not yet mature enough within AMD's drivers,
%which we plan to discuss with engineers at AMD.
%
We thus ran our experimental campaign on the Intel GPUs. The results
were similar across the GPUs, so for conciseness, we report only on
the Iris 6100 GPU (driver 20.19.15.4463) with a host CPU i3-5157U. The
Iris has a reported 47 compute units. Results for the two other Intel
GPUs are available in supplementary material.

@@

\myparagraph{Performance} \myfiglong\ref{fig:performance} reports the
overhead brought by the scheduling of non-cooperative kernels over the
cooperative kernel execution time.  This is the slowdown associated
with running the cooperative kernel in the presence of multitasking,
vs.\ running the cooperative kernel in isolation (median over all
applications and inputs).  We also show the period at which
non-cooperative kernels can be scheduled (median over all applications
and inputs). We use the median as our data included some drastic
outliers; these outliers appear to happen with applications/inputs in
which the resizing barrier are not called very frequently and the
graphics task asks for half or more resources. For example, a medium
graphics workload for bfs on the rmat input has over an 8$\times$
overhead, while the same application on the usa input has an overhead
of 1.11$\times$. As \myfiglong\ref{fig:performance} shows, most of our
applications/inputs are much better behaved than this. Future work
includes examining these problem applications/inputs in more detail,
possibly to insert more calls to $\offerfork$ and $\offerkill$.

@@

Let us assume that a graphics workload is required to be scheduled
with period $P$ and duration $D$, and that a long-running compute
kernel requires time $C$ to execute without interruption.  If we
assume the cost of preemption is negligible (e.g.\ Nvidia have
reported preemption times of 0.1 ms for Pascal~\cite{anandtech}), then
the overhead associated with switching between compute and graphics
every $P$ time steps is $P/(P-D)$.

We compare this task-level preemption overhead model with our
experimental results per graphics workload in
\mytab{\ref{tab:preemption}}. We report the overhead of the
configuration that allowed us to meet to deadline of the graphics
task.
%
Based on the above assumptions, our approach provides similar overhead
for low and medium graphics workloads; however, for the high workload,
our approach has a higher overhead.

@@

\myparagraph{GPU multitasking and scheduling}
%
%\ADComment{Tyler: please update or delete based on your recent readings and refer to Section~\ref{sec:nvidiacomparison}.}
%
Hardware support for preemption has been proposed for \nvidia GPUs, as
well as \emph{SM-draining} whereby workgroups occupying a symmetric
multiprocessor (SM; a compute unit using our terminology) are allowed
to complete until the SM becomes free for other
tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited
the presence of blocking constructs, since it may not be possible to
drain a blocked workgroup.
%
A follow-up work adds the notion of SM \emph{flushing}, where a
workgroup can be re-scheduled from scratch if it has not yet committed
side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have
been evaluated using simulators, over sets of regular GPU kernels.
Very recent \nvidia GPUs (i.e. the Pascal architecture) support
preemption, though, as discussted in \mysec{sec:intro} and
\mysec{sec:nvidiacomparison}, it is not clear whether they guarantee
fairness or allow tasks to share GPU resources at the workgroup
level~\cite{PascalWhitepaper}.

>>>>

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{overview.pdf}
\caption{Cooperative kernels can flexibly resize to let other tasks
(e.g. graphics) run concurrently}
\label{fig:overview}
\vspace{-2mm}
\end{figure}

@@

\myfiglong{\ref{fig:overview}} illustrates sharing of GPU compute units between a cooperative kernel and a
graphics task. Workgroups 2 and 3 of the cooperative kernel
are killed at an $\offerkill$ to make room for a graphics
task. The workgroups are subsequently restored to the cooperative kernel when workgroup 0 calls $\offerfork$. The \emph{gather time} is the time
between resources being requested and the application surrendering them via $\offerkill$. To satisfy soft-real time
constraints, this time should be low; our experimental study
(\mysec{\ref{sec:responsiveness}}) shows that in practice, the
gather-time for our applications is acceptable for a range of graphics
workloads.

@@

\begin{lstlisting}
(*@\label{line:wksteal:kernelfunc}@*)kernel work_stealing(global Task * queues) {
(*@\label{line:wksteal:getgroupid}@*)  int queue_id = get_group_id();
(*@\label{line:wksteal:mainloop}@*)  while (more_work(queues)) {
(*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
    if (t)
(*@\label{line:wksteal:processtask}@*)      process_task(t, queues, queue_id);
  }
}
\end{lstlisting}
\vspace{-2mm}
\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
%\vspace{-6mm}
\end{figure}

@@

\begin{lstlisting}
kernel graph_app(global graph * g,
                 global nodes * n0, global nodes * n1) {
  int level = 0;
  global nodes * in_nodes = n0;
  global nodes * out_nodes = n1;
  int tid = get_global_id();
  int stride = get_global_size();
(*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
    for (int i = tid; i < nodes.size; i += stride)
      process_node(g, in_nodes[i], out_nodes, level);
(*@\label{line:graph:swap}@*)    swap(&in_nodes, &out_nodes);
(*@\label{line:graph:gb1}@*)    global_barrier();
(*@\label{line:graph:reset}@*)    reset(out_nodes);
    level++;
(*@\label{line:graph:gb2}@*)    global_barrier();
  }
}
\end{lstlisting}
\vspace{-2mm}
\caption{An OpenCL graph traversal algorithm}\label{fig:graphsearch}
%\vspace{-2mm}
\end{figure}

@@

\begin{lstlisting}
kernel graph_app(global graph *g,
                 global nodes *n0, global nodes *n1) {
(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
  while(in_nodes.size > 0) {
(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
    for (int i = tid; i < nodes.size; i += stride)
      process_node(g, in_nodes[i], out_nodes, level);
    swap(&in_nodes, &out_nodes);
(*@\label{line:cgraph:resizing1}@*)    resizing_global_barrier();
    reset(out_nodes);
    level++;
(*@\label{line:cgraph:resizing2}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
%\vspace{-2mm}
\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
\vspace{-4mm}
\end{figure}

@@

On receiving a data packet describing a kernel launch $K$, the GPU
scheduler must decide how to schedule $K$. Suppose $K$ requests $N$
workgroups. The scheduler queries the scheduler context.  If there are
at least $N$ available workgroups, $K$ can be scheduled
immediately. Suppose instead that there are only $N_a < N$ available
workgroups, but a cooperative kernel $K_c$ is executing. The scheduler
can use $K_c$'s channel in the scheduler context to command $K_c$ to
provide $N - N_a$ workgroups via $\offerkill$.  Once $N$ workgroups
are available,
% the scheduler initialises a kernel context $\KC$ for $K$, setting
%the number of active workgroups to $N$ and mapping available
%workgroups (located in the scheduler context) to their new $\KC$
%workgroup ids.
the scheduler then sends $N$ workgroups from the available workgroups
to execute kernel $K$
% with
%kernel context $\KC$.
(If the new kernel $K$ is itself a cooperative kernel, the scheduler
would be free to provide $K$ with fewer than $N$ active workgroups
initially).

@@

\begin{table}[t]
\normalsize
\centering
\begin{tabular}{ l r r r r r r}
App. & barriers & kill & fork & transmit & LoC & inputs\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
%\rowcolor{Gray1}
%bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
\rowcolor{Gray3}
octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
\rowcolor{Gray3}
game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
\end{tabular} \\
%\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
\vspace{-10mm}
\end{table}

@@

Among the GPUs available to us, four met our requirements: Intel HD
520, HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux
drivers for OpenCL suffer from a known defect whereby long-running
kernels lead to unkillable defunct
processes~\cite{amd-defunct-process}.
%
%and simultaneously obtained alarming results on this platform when
%using SVM atomics: modifying a kernel to include a simple host/device
%handshake through SVM slowed down kernel execution by an order of
%magnitude in some cases.  This behaviour, combined with the defunct
%processes issue, lead us to suspect that the support we require for
%cooperative kernels is not yet mature enough within AMD's drivers,
%which we plan to discuss with engineers at AMD.
%
We thus ran our experiments on the Intel GPUs. The results
were similar across the GPUs, so for conciseness, we report only on
the Iris 6100 GPU (driver 20.19.15.4463) with a host CPU i3-5157U. The
Iris has a reported 47 compute units. Results for the other Intel GPUs
are available in supplementary material.

@@

\myparagraph{Performance} \myfiglong\ref{fig:performance} reports the
overhead brought by the scheduling of non-cooperative kernels over the
cooperative kernel execution time.  This is the slowdown associated
with running the cooperative kernel in the presence of multitasking,
vs.\ running the cooperative kernel in isolation (median over all
applications and inputs).  We also show the period at which
non-cooperative kernels can be scheduled (median over all applications
and inputs). We use the median as our data included some drastic
outliers; these outliers appear to happen with applications/inputs in
which the resizing barrier are not called very frequently and the
graphics task asks for half or more of the resources. For example, a
medium graphics workload for bfs on the rmat input has over an
8$\times$ overhead when asking for all but one of the workgroups for
graphics. As \myfiglong\ref{fig:performance} shows, most of our
applications/inputs are much better behaved than this. Future work
includes examining these problem applications/inputs in more detail,
possibly to insert more calls to $\offerfork$ and $\offerkill$.

@@

Given a graphics workload that is required to be scheduled with period
$P$ and duration $D$, and that a compute kernel requires time $C$ to
execute without interruption.  If we assume the cost of preemption is
negligible (e.g.\ Nvidia have reported preemption times of 0.1 ms for
Pascal~\cite{anandtech}), then the overhead associated with switching
between compute and graphics every $P$ time steps is $P/(P-D)$.

We compare this task-level preemption overhead model with our
experimental results per graphics workload in
\mytab{\ref{tab:preemption}}. We report the overhead of the
configuration that allowed us to meet to deadline of the graphics
task.
%
Based on the above assumptions, our approach provides similar overhead
for low and medium graphics workloads, however, has a higher overhead for
the high workload.

@@

\myparagraph{GPU multitasking and scheduling}
%
%\ADComment{Tyler: please update or delete based on your recent readings and refer to Section~\ref{sec:nvidiacomparison}.}
%
Hardware support for preemption has been proposed for \nvidia GPUs, as
well as \emph{SM-draining} whereby workgroups occupying a symmetric
multiprocessor (SM; a compute unit using our terminology) are allowed
to complete until the SM becomes free for other
tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited
the presence of blocking constructs, since it may not be possible to
drain a blocked workgroup.
%
A follow-up work adds the notion of SM \emph{flushing}, where a
workgroup can be re-scheduled from scratch if it has not yet committed
side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have
been evaluated using simulators, over sets of regular GPU kernels.
Very recent \nvidia GPUs (i.e. the Pascal architecture) support
preemption, though, as discussted in \mysec{\ref{sec:intro}} and
\mysec{\ref{sec:nvidiacomparison}}, it is not clear whether they guarantee
fairness or allow tasks to share GPU resources at the workgroup
level~\cite{PascalWhitepaper}.
